---
title: "Linear Mixed Models with Censoring Examples"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linear Mixed Models with Censoring Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, message=FALSE}
library(dplyr)

devtools::dev_mode(on = TRUE)
## knitr package options
knitr::opts_knit$set(verbose = FALSE)
options(digits = 3L)

library(survival)

library(lme4cens)
library(microbenchmark)

library(censReg)
library(crch)

data("Affairs", package = "lme4cens")

VERBOSE <- 0L
```


## Mixed Linear Models with censored observations

We use a censored version of the `sleepstudy`-data, by applying the information in column `event3` for left- and right-censored measurements of reaction time.
We use fixed boundaries (column `event3`) for left- and right-censoring so that we can compare the results with `censReg`.

```{r ex2_sleepstudy2_lme4cens}
REACT_L <- 212
REACT_R <- 350

sleepstudy2 %>% 
  lFormula(Surv(pmin(pmax(Reaction, REACT_L), REACT_R), time2 = Reaction, event = event3,
                type = "interval") ~ Days + (1|Subject), data = ., REML=FALSE) -> 
  lForm



lForm %>% 
  append(list(verbose=VERBOSE, quadrature = "stats")) %>%
  do.call(mkLmerCensDevfun_rInt_R, .) ->
  
  myDevFun_f

# Gauss-Hermite quadrature
lForm %>% 
  append(list(verbose=VERBOSE, quadrature = "gh")) %>%
  do.call(mkLmerCensDevfun_rInt_R, .) ->
  
  myDevFun_gh_f


## optimize function
paramS <- c(260, 5, 3, 2)

paramEst <- optim(par = paramS, fn = myDevFun_gh_f)
paramEst
```



### Comparison with `censReg`

We compare it with the results of `censReg` on the same data.

```{r ex2_sleepstudy2_censReg, message = FALSE}

sleepstudy2 %>% 
  dplyr::mutate_(Reaction = ~ pmin(pmax(Reaction, REACT_L), REACT_R))  %>% 
  plm::pdata.frame(index = c("Subject", "Days")) ->
  
  sleepstudy2.cr

fm.censReg <- censReg::censReg(Reaction ~ as.numeric(Days), left = REACT_L, right = REACT_R,
                               start = paramS, nGHQ = 8L, data = sleepstudy2.cr)

summary(fm.censReg)
```

The fitted coefficients are different, in particular the intercept coefficient ${\hat β}_0 =$ `r paramEst$par[1]` (`lme4cens`) vs `r coef(fm.censReg)[1]` (`censReg`). The mean value of the observations per time point are:
```{r sleepstudy2_desc, echo = FALSE, results='asis'}
sleepstudy2 %>%
  dplyr::group_by_(~ Days) %>%
  dplyr::summarise_(mReaction = ~ mean(Reaction)) %>% 
  knitr::kable(digits = 0L)
```




We use our method with start values that are close to the optimal values of the `censReg`-fit:
```{r ex2_sleepstudy2_lme4censLL}
optim(par = coef(fm.censReg), fn = myDevFun_gh_f)
```

The variance components change in comparison to the given initial optimal `censReg` values.

```{r ex2_sleepstudy2_censRegLL, echo = FALSE}
ll.censReg <- censReg::censReg(Reaction ~ as.numeric(Days), left = REACT_L, right = REACT_R, start = paramEst$par, logLikOnly = TRUE, nGHQ = 8L,
                               data = sleepstudy2.cr)

nll.lme4cens <- myDevFun_gh_f(param = paramEst$par)
nll.lme4cens.numIntStats <- myDevFun_f(param = paramEst$par)
```

Vice versa, we check the log-likelihood at the _optimal fit_ of our own implementation.
`censReg` calculates the log-likelihood contributions per subject, together with its gradient.
The log-likelihood contributions are `r paste(round(ll.censReg, 2), collapse = ", ")` which sum to `r sum(ll.censReg)`.
Our implementation using GH-quadrature at the same start vector gives a log-likelihood of `r -nll.lme4cens` with subject-contributions `r paste(round(log(attr(nll.lme4cens, "lik.contribs")), 2), collapse = ", ")`.
When using numeric integration from `stats`-package we get :
`r -nll.lme4cens.numIntStats` with subject-contributions `r paste(round(log(attr(nll.lme4cens.numIntStats, "lik.contribs")), 2), collapse = ", ")`




