---
title: "Linear Mixed Models with Censoring Examples"
author: "MK"
date: "2017-02-27, `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linear Mixed Models with Censoring Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE, message=FALSE}
library(dplyr)

devtools::dev_mode(on = TRUE)
## knitr package options
knitr::opts_knit$set(verbose = FALSE)
options(digits = 3L)

library(survival)

library(lme4cens)
library(microbenchmark)

library(censReg)
library(crch)

data("sleepstudy2", package = "lme4cens")

VERBOSE <- 0L
```


## Mixed Linear Models with censored observations

We use a censored version of the `sleepstudy`-data, by applying the information in column `event3` for left- and right-censored measurements of reaction time.
The response with these boundaries is stored in column `Reaction3`.
We use fixed boundaries (column `event3`) for left- and right-censoring so that we can compare the results with `censReg`.



The function `lmercens` is modelled after `lme4`'s `lmer` function.
One can specify directly own start values. Otherwise, `lmercens` chooses start values based on a `lmer` fit where censored data has been imputed.


### Own start values
We start with given start values for the parameters that stem from a ordinary `lmer`-fit on the underlying non-censored data.

```{r ex2_sleepstudy2_lme4cens1}
paramS <- c(260, 5, 3, 2)

fm.lmercens <- lmercens(Surv(Reaction3, time2 = Reaction3, event = event3, type = "interval") ~ Days + (1|Subject), data = sleepstudy2, 
              start = paramS, REML = FALSE)

fm.lmercens
```


### Automatic start values
```{r ex2_sleepstudy2_lme4cens2}
lmercens(Surv(Reaction3, time2 = Reaction3, event = event3, type = "interval") ~ Days + (1|Subject), data = sleepstudy2, 
              REML = FALSE)
```



### Comparison with `censReg`

We compare it with the results of `censReg` on the same data.

```{r ex2_sleepstudy2_censReg, message = FALSE}

REACT_L3 <- attr(sleepstudy2, "left3")
REACT_R3 <- attr(sleepstudy2, "right3")

sleepstudy2 %>% 
  plm::pdata.frame(index = c("Subject", "Days")) ->
  
  sleepstudy2.cr

fm.censReg <- censReg::censReg(Reaction3 ~ as.numeric(Days), left = REACT_L3, right = REACT_R3,
                               start = paramS, nGHQ = 8L, data = sleepstudy2.cr)

summary(fm.censReg)
```

The fitted coefficients are different, in particular the intercept coefficient ${\hat β}_0 =$ `r fm.lmercens$par[1]` (`lme4cens`) vs `r coef(fm.censReg)[1]` (`censReg`).
The mean value of the observations per time point are:

```{r sleepstudy2_desc, echo = FALSE, results='asis'}
sleepstudy2 %>%
  dplyr::group_by_(~ Days) %>%
  dplyr::summarise_(mReaction = ~ mean(Reaction3)) %>% 
  knitr::kable(digits = 0L)
```




We use our method with start values that are close to the optimal values of the `censReg`-fit:
```{r ex2_sleepstudy2_lme4censLL, echo = FALSE}
lmercens(Surv(Reaction3, time2 = Reaction3, event = event3, type = "interval") ~ Days + (1|Subject),
         start = coef(fm.censReg), data = sleepstudy2, 
         REML = FALSE)

fm.lmercens.devfun <- lmercens(Surv(Reaction3, time2 = Reaction3, event = event3, type = "interval") ~ Days + (1|Subject),
         start = coef(fm.censReg), data = sleepstudy2, 
         REML = FALSE, devFunOnly = TRUE)
```

The variance components change in comparison to the given initial optimal `censReg` values.

```{r ex2_sleepstudy2_censRegLL, echo = FALSE}
ll.censReg <- censReg::censReg(Reaction3 ~ as.numeric(Days), left = REACT_L3, right = REACT_R3, start = fm.lmercens$par, logLikOnly = TRUE, nGHQ = 8L,
                               data = sleepstudy2.cr)

nll.lme4cens <- fm.lmercens.devfun(param = fm.lmercens$par)
```

Vice versa, we check the log-likelihood at the _optimal fit_ of our own implementation.
`censReg` calculates the log-likelihood contributions per subject, together with its gradient.
The log-likelihood contributions are `r paste(round(ll.censReg, 2), collapse = ", ")` which sum to `r sum(ll.censReg)`.
Our implementation using GH-quadrature at the same start vector gives a log-likelihood of `r -nll.lme4cens` with subject-contributions `r paste(round(log(attr(nll.lme4cens, "lik.contribs")), 2), collapse = ", ")`.




