---
title: "Linear Mixed Models with Censoring Examples"
author: "Matthias Kuhn"
date: "2017-02-27, last update: `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linear Mixed Models with Censoring Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE, message=FALSE}

library('dplyr')
library('knitr')

## knitr package options
knitr::opts_knit$set(verbose = FALSE)
options(digits = 5L)

library('survival')

library('microbenchmark')

library('censReg')
library('crch')

library('lme4cens')


data("sleepstudy2", package = "lme4cens")

# verbosity-level
VERBOSE <- 0L

# manual start values:
# theta for random effect components: 1/ between and 2/ within
paramSL <- list(theta = c(3, 2), fixef = c(260, 5))
```



## Mixed Linear Models with censored observations

The function `lmercens` is modelled after `lme4`'s `lmer` function.
One can specify directly own start values. Otherwise, `lmercens` chooses start values based on a `lmer` fit where censored data had been imputed.


### sleepstudy2 data
As an example dataset we use a censored version of the well-known `sleepstudy`-data,
by applying the __fixed boundary left- and right-censoring__ (see column `event3`) on the measurements of reaction time.
The response with these boundaries is stored in column `Reaction3`.
Because of fixed boundaries for censoring we can compare our results with those from the package `censReg` (that allows only for fixed censoring boundaries).

We use a subset of `r NROW(sleepstudy2)` observations from `r nlevels(sleepstudy2$Subject)` subjects of the original `sleepstudy`-dataset.
The mean value of the observations per time point are:

```{r sleepstudy2_desc, echo = VERBOSE > 0, results='asis'}
sleepstudy2 %>%
  dplyr::group_by(Days) %>%
  dplyr::summarise(`mean~Reaction~` = mean(Reaction3), `median~Reaction~` = median(Reaction3)) %>% 
  knitr::kable(digits = 0L)
```



### Own start values
We manually provide start values from an ordinary `lmer`-fit on the underlying non-censored data to the `lmercens`-call.
Start values — stored in the list `paramSL` — are:

* `r paramSL$theta` for random effects (between and within std. deviation on log-scale) _and_
* `r paramSL$fixef` for fixed effects

The model fit is: 

```{r ex2_sleepstudy2_lme4cens1, echo = TRUE}
fm.lmercens <- lmercens(Surv(Reaction3, time2 = Reaction3, event = event3, type = "interval") ~ Days + (1|Subject),
                        data = sleepstudy2, REML = FALSE,
                        control = lmerControl(optimizer = "optimx", optCtrl = list(method = "L-BFGS-B"), boundary.tol = 0),
                        start = paramSL) 
summary(fm.lmercens)
```

We can request a confidence-interval for the fixed effect parameters. It is based on asymptotic ML-theory and uses the observed Fisher-information matrix to get an estimate of the standard errors for the parameters.

```{r ex2_sleepstudy2_lme4cens_confint, echo = TRUE}
confint(fm.lmercens, level = 0.95)
```

The predictions for the random effects is extracted via `ranef` (like in the `lme4` package):
```{r ex2_sleepstudy2_lme4cens_ranef, echo = TRUE}
ranef(fm.lmercens)
```

Model predictions can be made on the training data either with or without the random effect predictions:
```{r ex2_sleepstudy2_lme4cens_predict, echo = TRUE}
cbind(get_all_vars(fm.lmercens, data = sleepstudy2),
      pred_wRanEff = predict(fm.lmercens),
      pred_fixedOnly = predict(fm.lmercens, re.form = ~0))
```


### Automatic start values
If we omit the start values `lmercens` will first do a call to `lmer` where the censored observations have been imputed.

```{r ex2_sleepstudy2_lme4cens2, echo = VERBOSE > 0}
lmercens(Surv(Reaction3, time2 = Reaction3, event = event3, type = "interval") ~ Days + (1|Subject),
         data = sleepstudy2, REML = FALSE,
         control = lmerControl(optimizer = "optimx", optCtrl = list(method = "L-BFGS-B"), boundary.tol = 0)) %>% 
  summary
```

The result (at least for the intercept term) depends on the start values. Is there a local minimum where we are stuck in the optimization?




### Comparison with `censReg`

With the package `censReg` you can fit linear mixed models where the metric response variable can be left- and/or right-censored at __fixed__ boundary values.
We compare our results here with the results of `censReg` on the same data and using fixed start vector.

```{r ex2_sleepstudy2_censReg, echo = FALSE, message = FALSE}

REACT_L3 <- attr(sleepstudy2, "left3")
REACT_R3 <- attr(sleepstudy2, "right3")

sleepstudy2 %>% 
  plm::pdata.frame(index = c("Subject", "Days")) ->
  
  sleepstudy2.cr

fm.censReg <- censReg::censReg(Reaction3 ~ as.numeric(Days), left = REACT_L3, right = REACT_R3,
                               start = unlist(rev(paramSL)), nGHQ = 8L, data = sleepstudy2.cr)

summary(fm.censReg)

# likelihood contributions at censReg's optimal parameters calculated by censReg
ll_censReg <- censReg::censReg(Reaction3 ~ as.numeric(Days), left = REACT_L3, right = REACT_R3,
                               start = coef(fm.censReg), nGHQ = 8L, data = sleepstudy2.cr,
                               logLikOnly = TRUE)
```

The fitted coefficients are different, in particular the fixed effect intercept coefficient:

* ${\hat β}_0 =$ `r fixef(fm.lmercens)[1]` (`lme4cens`)
* ${\hat β}_0 =$ `r coef(fm.censReg)[1]` (`censReg`)




#### At `censReg`'s optimal fit
We evaluate our fit  at the optimal values from the `censReg`-fit (i.e. no further optimization).
We look at the contribution to the log-likelihood per subject and compare the two fitting routines.
`censReg` calculates the log-likelihood contributions per subject, together with its gradient.

```{r ex2_sleepstudy2_lme4censLL, echo = FALSE}
fm.lmercens.devfun <- lmercens(Surv(Reaction3, time2 = Reaction3, event = event3, type = "interval") ~ Days + (1|Subject),
         data = sleepstudy2, REML = FALSE, devFunOnly = TRUE)

ll_usOnUs <- fm.lmercens.devfun(param = fm.lmercens$par)
ll_usOnCensReg <- fm.lmercens.devfun(coef(fm.censReg)[c(3,4,1,2)])
```

```{r ex2_sleepstudy2_LL_optimCensReg, echo = FALSE, results='asis'}
tibble::tibble(source = c(paste("Subject", levels(sleepstudy2$Subject)), "total"), 
               censReg = c(as.numeric(ll_censReg), sum(ll_censReg)),
               lmercens = c(log(attr(ll_usOnCensReg, which = "lik.contribs")), -ll_usOnCensReg),
               `Δ~lmercens-censReg~` = lmercens - censReg) %>% 
  knitr::kable(caption = "Log-likelihood contributions at __censReg__'s optimal values on sleepstudy2-data", digits = 3)
```

Differences between `censReg` and `lmercens` are small and are probably due to numerical issues.



#### At `lmercens`'s optimal fit
Vice versa, we compare the log-likelihood contributions at the _optimal fit_ of `lmercens`.

```{r ex2_sleepstudy2_censRegLL, echo = FALSE}
ll_censRegOnUs <- censReg::censReg(Reaction3 ~ as.numeric(Days), left = REACT_L3, right = REACT_R3,
    start = fm.lmercens$par[c(3,4,1,2)], logLikOnly = TRUE, nGHQ = 8L,
    data = sleepstudy2.cr)
```



```{r ex2_sleepstudy2_LL_optimLmercens, echo = FALSE, results='asis'}
dplyr::tibble(source = c(paste("Subject", levels(sleepstudy2$Subject)), "total"), 
               censReg = c(as.numeric(ll_censRegOnUs), sum(ll_censRegOnUs)),
               lmercens = c(log(attr(ll_usOnUs, which = "lik.contribs")), -ll_usOnUs),
               `Δ~lmercens-censReg~` = lmercens - censReg) %>%
  knitr::kable(caption = "Log-likelihood contributions at __lmercens__'s optimal values on sleepstudy2-data")
```

The log-likelihood contributions of all subjects as estimated by `lmercens` are consistantly bigger (less negative) than by `censReg`.



### Comparison with `survreg`
The function `survreg` does not support for random effects,
although a `frailty()` term within the formula used to work (until `survival` version 3.2-8).
But, also before it did not work properly, citing from the `NEWS`-file of package `survival`:

> it is integrating out the appropriate density for a coxph model, which is the wrong one for `survreg`


