---
title: "Example analyses of simple linear mixed models"
author: "MK"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Example analysis of simple linear mixed models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, message=FALSE}
## knitr package options
knitr::opts_knit$set(verbose = FALSE)
options(digits = 3L)


library(dplyr)
devtools::dev_mode(on = TRUE)

library(lme4cens)

data("sleepstudy2", package = "lme4cens")
data("Pastes", package = "lme4")

VERBOSE <- 0L

DO_REML <- FALSE
```


We test our linear mixed model R-routines with some small examples and compare the fit with the fit of the official `lme4` implementation.
We use a subset of the infamous `sleepstudy`-data to fit simple linear mixed models.

Throughout, we use `r if (DO_REML) "__REML__" else "__ML__"` for model fitting.

If no censoring occurs, the optimization function (e.g. log-likelihood for ML) is parameterized by
the variance parameters θ for the random effects.
The fixed effect estimates and residual variance follow, given the variance parameters θ.

In case of censoring I do not see an easy way to also profile out these parameters.



## Re-Implementation following `lme4`

### Random intercept model

The single random parameter θ corresponds to the root of the scaled between-subject variance.
Here, we examplify how our objective function (implemented in R) is created and the optimization then started manually.

```{r ex1_lmm_ranInt_sleepstudy2}

lFormula(Reaction ~ Days + (1|Subject), data = sleepstudy2, REML=DO_REML) %>% 
  append(list(verbose=VERBOSE)) %>% 
  do.call(mkLmerDevfun_R, .) ->
  
  myDevFun_f

# one-dimensional minimization on theta-parameter
optRes <- optimize(myDevFun_f, interval = c(0.01, 15))

(theta_h <- optRes$minimum)
attributes(optRes$objective)
```

The optimization routine gives parameter estimate $\hat{θ} =$ `r theta_h`.
The squared estimate is the estimated scaled between-subject variance, $\hat{θ}^2 =$  `r theta_h^2`.
The unscaled between-subject variance is hence estimated as $\hat{σ}^2 · \hat{θ}^2 =$ `r attr(optRes$objective, "resVar") * theta_h^2`.


The estimates match those of `lme4`:
```{r ex1_lmm_ranInt_sleepstudy2_lme4}
summary(fm <- lmer(Reaction ~ Days + (1|Subject), data = sleepstudy2,
                   REML=DO_REML))
ranef(fm)
```







### Random Intercept With weighting
We apply an arbitary weighting to the observations.

```{r ex2_lmm_ranInt_w_sleepstudy2}
w_sleepstudy2 <- rep(c(rep(2,3), rep(1,7)), 4)
stopifnot( length(w_sleepstudy2) == NROW(sleepstudy2) )

lFormula(Reaction ~ Days + (1|Subject), REML = DO_REML,
         weights = w_sleepstudy2, data = sleepstudy2) %>% 
  append(values = list(verbose=VERBOSE)) %>% 
  do.call(mkLmerDevfun_R, .) ->
  
  myDevFun_fw


optRes <- optimize(myDevFun_fw, interval = c(0.01, 19))
theta_h <- optRes$minimum

attributes(optRes$objective)
```

The optimization routine gives $\hat θ =$ `r theta_h` which corresponds to an estimated scaled between-subject variance of `r theta_h**2`.
The unscaled between-subject variance is estimated as `r theta_h^2 * attr(optRes$objective, "resVar")`.

The results match the results from `lme4`:
```{r ex2_lmm_ranInt_w_sleepstudy2_lme4}
fm <- lmer(Reaction ~ Days + (1|Subject),
           weights = w_sleepstudy2, 
           data = sleepstudy2, REML=DO_REML)
summary(fm)
ranef(fm)
```




### Random intercept-slope model

We have $k=1$ random-effect factor with $l_1 = 4$ levels and $p_1 = 2$ columns (intercept and `Days`).
Here, $θ$ is a three-dimensional parameter.

```{r ex1_lmm_ranIntSlo_sleepstudy2}
lFormula(Reaction ~ Days + (Days|Subject), data = sleepstudy2, REML=DO_REML) %>% 
  append(values = list(verbose=VERBOSE)) %>% 
  do.call(mkLmerDevfun_R, .) ->
  
  myDevFun_f

# three-dimensional minimization of theta-parameter
optRes <- optim(par = c(1, 0, 1), fn = myDevFun_f, method = "BFGS")

theta_h <- optRes$par

modelParamEstim <- attributes(myDevFun_f(theta = theta_h))
modelParamEstim
```

The optimization routine gives $θ = ( `r theta_h` )$ which builds to the relative covariance factor:
```{r ex1_lmm_ranIntSlo_sleepstudy2_relCoVar, echo=FALSE, results='markup'}
nc <- 2
rowIndices <- rep(1:nc, 1:nc)
colIndices <- sequence(1:nc)
relCovarFactor <- Matrix::sparseMatrix(i = rowIndices, j = colIndices,
                                    x = as.numeric(rowIndices == colIndices) )
relCovarFactor@x <- theta_h
relCovarFactor
```

The estimates on the diagonal of the relative covariance factor are by convention taken as absolute value,
the sign does not matter for the resulting estimated covariance-matrix $Σ_θ$ of the random-effects vector B.
It is $Σ_θ = σ^2 Λ_θ Λ_θ^T$.

```{r ex1_lmm_ranIntSlo_sleepstudy2_REcovar}
modelParamEstim[["resVar"]] * Matrix::tcrossprod(relCovarFactor)
cov2cor(modelParamEstim[["resVar"]] * Matrix::tcrossprod(relCovarFactor))
```


We compare it with the output of `lme4`:
```{r ex1_lmm_ranIntSlo_sleepstudy2_lme4}
summary(fm <- lmer(Reaction ~ Days + (Days|Subject), data = sleepstudy2, REML=DO_REML))
ranef(fm)
```




### Crossed random effects

We use the `Penicillin`-data from `lme4`-package. The effect of six different penicillin samples is measured as the `diameter` of the zone of inhibition of bacteria growth. Six measurements are done on a single plate that acts like a blocking factor.
We have a randomized complete block design (RCBD), i.e. each sample is measured exactly once on each plate.

```{r ex2_lmm_crossed_Penicillin} 
data("Penicillin", package = "lme4")
lFormula(diameter ~ (1|plate) + (1|sample), data = Penicillin, REML = DO_REML) %>% 
  append(list(verbose=VERBOSE)) %>% 
  do.call(mkLmerDevfun_R, .) ->
  
  myDevFun_f


optRes <- optim(par = c(1, 2), fn = myDevFun_f)

theta_h <- optRes$par

# evaluate the objective-function at the optimal value
modelParamEstim <- attributes(myDevFun_f(theta = theta_h))
modelParamEstim
```

The optimization routine gives $θ = ( `r theta_h` )$ which corresponds to two scaled between-subject variance of $(`r theta_h**2`)$.
The unscaled between-subject variances for the two random effects are estimated as $(`r theta_h^2 * modelParamEstim$resVar`)$.


We compare it with the standard `lme4`-output of the same model specification:
```{r ex2_lmm_crossed_Penicillin_lme4}
summary(fm <- lmer(diameter ~ (1|plate) + (1|sample), data = Penicillin, REML = DO_REML))
ranef(fm)
```






### Nested random effects

We use a dataset with a response (`strength`) which depends on 
$k=2$ different factors with simple ($p_1 = p_2 = 1$) random effects:

* batch effect
* cask effect within batch

Hence, $θ$ consists of two parameters.
```{r ex3_lmm_nested_Pastes}
lFormula(strength ~ (1|batch/cask), data = Pastes, REML=DO_REML) %>% 
  append(values = list(verbose=VERBOSE)) %>% 
  do.call(mkLmerDevfun_R, .) ->
  
  myDevFun_f


optRes <- optim(par = c(1, 1), fn = myDevFun_f)

theta_h <- optRes$par

modelParamEstim <- attributes(myDevFun_f(theta = theta_h))
modelParamEstim
```

The optimization yields $\hat θ = ( `r theta_h`)$ 
which corresponds to two scaled between-subject variance of $(`r theta_h**2`)$.

The unscaled between-subject variances for the two random effects are estimated as $(`r theta_h^2 * modelParamEstim$resVar`)$.
The last ten random effect predictions are for the batch effect.
I guess the `batch:cask` factor comes first as it has more levels.

We compare it with the `lme4`-result:
```{r ex3_lmm_nested_Pastes_lme4}
summary(fm <- lmer(strength ~ (1|batch/cask), data = Pastes, REML=DO_REML))
ranef(fm)
```






## Own Implementation (allowing for censoring)

We consider again the simplest case of a linear mixed effect model, the simple random intercept model on the `sleepstudy2`-data.
With `lme4` we had:

```{r ref.label="ex1_lmm_ranInt_sleepstudy2_lme4", results='markup'}
```


We use the same data -- no censoring in the response variable.
But this time we send it through our own (direct integration) implementation for the likelihood that supports censoring.
Concretely, we wrap the response into `Surv` (with no status variable as we have no censoring in the dataset)
and call the function `lmercens` that internally uses `mkLmerCensDevfun_rInt_R` to build the optimization function.

One big difference to the standard implementation (without censoring) is that we optimize _four_ parameters instead of a single _one_, namely

* two variance parameters σ~bw~ and σ~res~ on log scale _and_
* two fixed effect parameters β~int~ and β~slope~.

Also, the between-subject parameter σ~bw~ is un-scaled, i.e. _not_ scaled by the residual variance.

```{r ex1_lmm_ranInt_sleepstudy2_cens, warning=FALSE, echo = FALSE, results='asis'}
library(survival)
library(optimx)

parNames <- c("ln(σ~bw~)", "ln(σ~res~)", "β~int~", "β~slo~")
startL <- list(theta = c(3, 2), fixef = c(260, 5))

# Gauss-Hermite quadrature with Nelder-Mead optimizer
fm_nm <- lmercens(Surv(Reaction) ~ Days + (1|Subject), data = sleepstudy2, REML=DO_REML,
         start = startL, control = lmerControl(optimizer = "Nelder_Mead"))

# Gauss-Hermite quadrature with bobyqa optimizer
fm_bo <- lmercens(Surv(Reaction) ~ Days + (1|Subject), data = sleepstudy2, REML=DO_REML,
         start = startL, control = lmerControl(optimizer = "bobyqa"))

# Gauss-Hermite quadrature with BFGS
fm_bfgs <- lmercens(Surv(Reaction) ~ Days + (1|Subject), data = sleepstudy, REML=DO_REML,
                    start = startL, control = lmerControl(optimizer = "optimx", optCtrl = list(method = "L-BFGS-B")))


# display results
dplyr::inner_join(
  x=dplyr::inner_join(
    x= c(fm_nm$par, -fm_nm$fval, exp(2*fm_nm$par[1]), exp(2*fm_nm$par[2])) %>% 
      setNames(nm = c(parNames, "Log-Likelihood", "σ~bw~^2^", "σ~res~^2^")) %>% 
      tibble::enframe(name = "Parameter", value = "NM"),
    
    y= c(fm_bo$par, -fm_bo$fval, exp(2*fm_bo$par[1]), exp(2*fm_bo$par[2])) %>% 
      setNames(nm = c(parNames, "Log-Likelihood", "σ~bw~^2^", "σ~res~^2^")) %>% 
      tibble::enframe(name = "Parameter", value = "bobyqa"),
    
    by = "Parameter"
  ),
  y=c(fm_bfgs$par, -fm_bfgs$fval, exp(2*fm_bfgs$par[1]), exp(2*fm_bfgs$par[2])) %>% 
      setNames(nm = c(parNames, "Log-Likelihood", "σ~bw~^2^", "σ~res~^2^")) %>% 
      tibble::enframe(name = "Parameter", value = "L-BFGS-B"),
  
  by = "Parameter"
)

```


Neither optimization methods meet the `lme4`-fit, in particular `L-BFGS-B` is off-target
and the other gradient-free methods deliver different estimates for the variance components.
The `NM`-method comes closer to the `lme4`-fit than the `bobyqa`-method.
Note that NM-method yields the biggest log-likelihood value (i.e. least negative),
even bigger than the `lme4`-fit.


We use the NM-fit for parameter θ and plug it into `lmer` to check the log-likelihood (if also `lme4` confirms this better log-likelihood).
The NM-based estimate for scaled between-subject variability as standard deviation is 
$\hat{θ}_{\text{NM}}=$ `r (theta_nm <- exp(fm_nm$par[1] - fm_nm$par[2]))`.

```{r ex1_lmm_ranInt_sleepstudy2_lme4_w_start_from_cens, echo = TRUE}
lmer( Reaction ~ Days + (1|Subject), data = sleepstudy2, REML = DO_REML,
      start = theta_nm, control = lmerControl(optimizer=NULL)) %>% 
  summary
```
The parameter fits returned by `lme4` are not identical to those of our censoring-aware implementation.
Package `lme4` does not confirm our log-likelihood value but the log-likelihood is not changed compared to the `lme4`-fit which hints that the log-likelihood function is very flat in the neighbourhood of the MLE.



