---
title: "Example analyses of simple linear mixed models"
author: "MK"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
vignette: >
  %\VignetteIndexEntry{Example: sleepstudy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, message=FALSE}

library(dplyr)
devtools::dev_mode(on = TRUE)
## knitr package options
knitr::opts_knit$set(verbose = FALSE)
options(digits = 3L)

library(lme4cens)

data("sleepstudy2", package = "lme4cens")
data("Pastes", package = "lme4")

VERBOSE <- 0L

DO_REML <- FALSE
```


We test our linear mixed model R-routines with some small examples and compare the fit with the fit of the official `lme4` implementation.
We use a subset of the infamous `sleepstudy`-data to fit simple linear mixed models.



### Random intercept model

The single random parameter θ corresponds to the root of the scaled between-subject variance.
Here, we examplify how our objective function (implemented in R) is created and the optimization then started manually.

```{r ex1_lmm_ranInt_sleepstudy2}

lFormula(Reaction ~ Days + (1|Subject), data = sleepstudy2, REML=DO_REML) %>% 
  append(list(verbose=VERBOSE)) %>% 
  do.call(mkLmerDevfun_R, .) ->
  
  myDevFun_f

# one-dimensional minimization on theta-parameter
optRes <- optimize(myDevFun_f, interval = c(0.01, 15))

theta_h <- optRes$minimum
attributes(optRes$objective)
```

The optimization routine gives parameter estimate θ = `r theta_h` which corresponds to a scaled between-subject variance of `r theta_h^2`.
The unscaled between-subject variance is estimated as `r theta_h^2 * attr(optRes$objective, "resVar")`.


We compare it with the output of `lme4`:
```{r ex1_lmm_ranInt_sleepstudy2_lme4}
summary(fm <- lmer(Reaction ~ Days + (1|Subject), data = sleepstudy2,
                   REML=DO_REML))
ranef(fm)
```


#### LmerCens_rInt function
We use the same data -- no censoring in the response variable.
But this time we send it through our own (direct integration) implementation for the likelihood that supports censoring.

```{r ex1_lmm_ranInt_sleepstudy2_cens}
library(survival)

# Gauss-Hermite quadrature
fm <- lmercens(Surv(Reaction) ~ Days + (1|Subject), data = sleepstudy2, REML=DO_REML,
         start = list(theta = c(3, 2), fixef = c(260, 5)), control = lmerControl(optimizer = "Nelder_Mead"))

fm$par
```


The fixed effect parameters that come out are

* intercept β~0~ = `r round(fm$par[1], 4L)` _and_ 
* slope β~1~ = `r round(fm$par[2], 4L)`.

The between-subject variance is estimated as `r sprintf("%4.f", exp(2*fm$par[1]))`
while the residual variance as `r sprintf("%4.f", exp(2*fm$par[2]))`.
The likelihood is `r -fm$fval`.





#### With weighting
We apply an arbitary weighting to the observations.

```{r ex2_lmm_ranInt_w_sleepstudy2}
w_sleepstudy2 <- rep(c(rep(2,3), rep(1,7)), 4)
stopifnot( length(w_sleepstudy2) == NROW(sleepstudy2) )

lFormula(Reaction ~ Days + (1|Subject),
         weights = w_sleepstudy2, data = sleepstudy2) %>% 
  append(values = list(verbose=VERBOSE)) %>% 
  do.call(mkLmerDevfun_R, .) ->
  
  myDevFun_fw


optRes <- optimize(myDevFun_fw, interval = c(0.01, 19))
theta_h <- optRes$minimum

attributes(optRes$objective)
```

The optimization routine gives $\theta = `r theta_h`$ which corresponds to a scaled between-subject variance of `r theta_h**2`.
The unscaled between-subject variance is estimated as `r theta_h^2 * attr(optRes$objective, "resVar")`.

We compare it with the output of `lme4`:
```{r ex2_lmm_ranInt_w_sleepstudy2_lme4}
fm <- lmer(Reaction ~ Days + (1|Subject),
            weights = w_sleepstudy2, 
             data = sleepstudy2, REML=T)
ranef(fm)
summary(fm)
```





### Random intercept-slope model

We have $k=1$ random-effect factor with $l_1 = 4$ levels and $p_1 = 2$ columns (intercept and `Days`).
Here, $\theta$ is a three-dimensional parameter.
```{r ex1_lmm_ranIntSlo_sleepstudy2}

lFormula(Reaction ~ Days + (Days|Subject), data = sleepstudy2, REML=T) %>% 
  append(values = list(verbose=VERBOSE)) %>% 
  do.call(mkLmerDevfun_R, .) ->
  
  myDevFun_f

# three-dimensional minimization of theta-parameter
optRes <- optim(par = c(1, 0, 1), fn = myDevFun_f)

theta_h <- optRes$par

modelParamEstim <- attributes(myDevFun_f(theta = theta_h))
modelParamEstim
```

The optimization routine gives $\theta = ( `r theta_h` )$ which builds to the relative covariance factor:
```{r ex1_lmm_ranIntSlo_sleepstudy2_relCoVar, echo=FALSE, results='markup'}
nc <- 2
rowIndices <- rep(1:nc, 1:nc)
colIndices <- sequence(1:nc)
relCovarFactor <- Matrix::sparseMatrix(i = rowIndices, j = colIndices,
                                    x = as.numeric(rowIndices == colIndices) )
relCovarFactor@x <- theta_h
relCovarFactor
```

The estimates on the diagonal of the relative covariance factor are by convention taken as absolute value,
the sign does not matter for the resulting estimated covariance-matrix $\Sigma_\theta$ of the random-effects vector B.
It is $\Sigma_\theta = \sigma^2 \Lambda_\theta \Lambda_\theta^T$.
```{r ex1_lmm_ranIntSlo_sleepstudy2_REcovar}
modelParamEstim[["resVar"]] * Matrix::tcrossprod(relCovarFactor)
cov2cor(modelParamEstim[["resVar"]] * Matrix::tcrossprod(relCovarFactor))
```


We compare it with the output of `lme4`:
```{r ex1_lmm_ranIntSlo_sleepstudy2_lme4}
summary(fm <- lmer(Reaction ~ Days + (Days|Subject),
                   data = sleepstudy2, REML=T))
ranef(fm)
```




### Crossed random effects

We use the `Penicillin`-data from `lme4`-package. The effect of six different penicillin samples is measured as the `diameter` of the zone of inhibition of bacteria growth. Six measurements are done on a single plate that acts like a blocking factor.
We have a randomized complete block design (RCBD), i.e. each sample is measured exactly once on each plate.

```{r ex2_lmm_crossed_Penicillin} 
data("Penicillin", package = "lme4")
lFormula(diameter ~ (1|plate) + (1|sample), data = Penicillin, REML = TRUE) %>% 
  append(list(verbose=VERBOSE)) %>% 
  do.call(mkLmerDevfun_R, .) ->
  
  myDevFun_f


optRes <- optim(par = c(1, 2), fn = myDevFun_f)

theta_h <- optRes$par

# evaluate the objective-function at the optimal value
modelParamEstim <- attributes(myDevFun_f(theta = theta_h))
modelParamEstim
```

The optimization routine gives $\theta = ( `r theta_h` )$ which corresponds to two scaled between-subject variance of $(`r theta_h**2`)$.
The unscaled between-subject variances for the two random effects are estimated as $(`r theta_h^2 * modelParamEstim$resVar`)$.


We compare it with the standard `lme4`-output of the same model specification:
```{r ex2_lmm_crossed_Penicillin_lme4}
summary(fm <- lmer(diameter ~ (1|plate) + (1|sample), data = Penicillin, REML = TRUE))
ranef(fm)
```






### Nested random effects

We use a dataset with a response (`strength`) which depends on 
$k=2$ different factors with simple ($p_1 = p_2 = 1$) random effects:

* batch effect
* cask effect within batch

Hence, $\theta$ consists of two parameters.
```{r ex3_lmm_nested_Pastes}
lFormula(strength ~ (1|batch/cask), data = Pastes, REML=T) %>% 
  append(values = list(verbose=VERBOSE)) %>% 
  do.call(mkLmerDevfun_R, .) ->
  
  myDevFun_f


optRes <- optim(par = c(1, 1), fn = myDevFun_f)

theta_h <- optRes$par

modelParamEstim <- attributes(myDevFun_f(theta = theta_h))
modelParamEstim
```
The last ten random effect predictions are for the batch effect.
I think the `batch:cask` factor comes first as it has more levels.

We compare it with the `lme4`-result:
```{r ex3_lmm_nested_Pastes_lme4}
summary(fm <- lmer(strength ~ (1|batch/cask), data = Pastes, REML=T))
ranef(fm)
```

