---
title: "Linear Mixed Models with Censoring Examples"
author: "MK"
date: "2017-02-27, `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linear Mixed Models with Censoring Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE, message=FALSE}
library(dplyr)

devtools::dev_mode(on = TRUE)
## knitr package options
knitr::opts_knit$set(verbose = FALSE)
options(digits = 3L)

library(survival)

library(lme4cens)
library(microbenchmark)

library(censReg)
library(crch)

data("sleepstudy2", package = "lme4cens")

# verbosity-level
VERBOSE <- 0L

# start value
paramS <- c(260, 5, 3, 2)
```



## Mixed Linear Models with censored observations

The function `lmercens` is modelled after `lme4`'s `lmer` function.
One can specify directly own start values. Otherwise, `lmercens` chooses start values based on a `lmer` fit where censored data had been imputed.


### sleepstudy2 data
For data we use a censored version of the `sleepstudy`-data, by applying the fixed boundary left- and right-censoring (column `event3`) on the measurements of reaction time.
The response with these boundaries is stored in column `Reaction3`. Because of fixed boundaries for censoring we can compare the results with `censReg`.

The mean value of the observations per time point are:

```{r sleepstudy2_desc, echo = VERBOSE > 0, results='asis'}
sleepstudy2 %>%
  dplyr::group_by_(~ Days) %>%
  dplyr::summarise_(mReaction = ~ mean(Reaction3)) %>% 
  knitr::kable(digits = 0L)
```



### Own start values
We start with given start values for the parameters that stem from an ordinary `lmer`-fit on the underlying non-censored data.

```{r ex2_sleepstudy2_lme4cens1, echo = VERBOSE > 0}

fm.lmercens <- lmercens(Surv(Reaction3, time2 = Reaction3, event = event3, type = "interval") ~ Days + (1|Subject),
                        data = sleepstudy2,
                        method = "BF", 
                        start = paramS, REML = FALSE) 

fm.lmercens
```


### Automatic start values
```{r ex2_sleepstudy2_lme4cens2}
lmercens(Surv(Reaction3, time2 = Reaction3, event = event3, type = "interval") ~ Days + (1|Subject),
          method = "BFGS",
         data = sleepstudy2, REML = FALSE)
```



### Comparison with `censReg`

We compare it with the results of `censReg` on the same data and using fixed start vector.

```{r ex2_sleepstudy2_censReg, message = FALSE}

REACT_L3 <- attr(sleepstudy2, "left3")
REACT_R3 <- attr(sleepstudy2, "right3")

sleepstudy2 %>% 
  plm::pdata.frame(index = c("Subject", "Days")) ->
  
  sleepstudy2.cr

fm.censReg <- censReg::censReg(Reaction3 ~ as.numeric(Days), left = REACT_L3, right = REACT_R3,
                               start = paramS, nGHQ = 8L, data = sleepstudy2.cr)

summary(fm.censReg)

# likelihood contributions at censReg's optimal parameters calculated by censReg
ll_censReg <- censReg::censReg(Reaction3 ~ as.numeric(Days), left = REACT_L3, right = REACT_R3,
                               start = coef(fm.censReg), nGHQ = 8L, data = sleepstudy2.cr, logLikOnly = TRUE)
```

The fitted coefficients are different, in particular the intercept coefficient ${\hat β}_0 =$ `r fm.lmercens$par[1]` (`lme4cens`) vs `r coef(fm.censReg)[1]` (`censReg`).
`censReg` gives as contributions per subject `r paste(round(as.vector(ll_censReg),3), collapse = ", ")`. 


#### Us on `censReg`'s fit
We use our method and evaluate it at the `censReg`-fit (i.e. no further optimization).

```{r ex2_sleepstudy2_lme4censLL, echo = FALSE}
fm.lmercens.devfun <- lmercens(Surv(Reaction3, time2 = Reaction3, event = event3, type = "interval") ~ Days + (1|Subject),
         start = coef(fm.censReg), 
         data = sleepstudy2, REML = FALSE, devFunOnly = TRUE)
ll.usOnCensReg <- fm.lmercens.devfun(coef(fm.censReg))
```

We calculate a log-likelihood of `r as.vector(-ll.usOnCensReg)`.
Our log-likelihood contributions at the `censReg`'s fitted parameter are `r paste(round(log(attr(ll.usOnCensReg, "lik.contribs")), 3), collapse = ", ")`.
They are very close to `censReg`'s output. Differences are probably due to numerical issues.



#### `censReg` on us
Vice versa, we check the log-likelihood at the _optimal fit_ of our own implementation.

```{r ex2_sleepstudy2_censRegLL, echo = FALSE}
ll.censRegOnUs <- censReg::censReg(Reaction3 ~ as.numeric(Days), left = REACT_L3, right = REACT_R3,
    start = fm.lmercens$par, logLikOnly = TRUE, nGHQ = 8L,
    data = sleepstudy2.cr)

ll.usOnUs <- fm.lmercens.devfun(param = fm.lmercens$par)
```

`censReg` calculates the log-likelihood contributions per subject, together with its gradient.
The log-likelihood contributions as calculated by `censReg` at our optimal parameter are `r paste(round(ll.censRegOnUs, 3), collapse = ", ")` which sum to `r sum(ll.censRegOnUs)`.
Our implementation using GH-quadrature at the same start vector gives a log-likelihood of `r -ll.usOnUs` with subject-contributions `r paste(round(log(attr(ll.usOnUs, "lik.contribs")), 3), collapse = ", ")`.




### Comparison with `survreg`
`survreg` has basic support for random effects, using the `frailty()` within the formula.
Here, we use Gaussian frailty with the default `reml`-method although
in the help-page to `frailty` the authors issue a warning that frailty in parametric models `survreg` the combinations `gamma/ml` and `gaussian/reml` do _not_ lead to correct results.


```{r ex2_sleepstudy2_survreg}
fm.survreg <- survreg(Surv(Reaction3, time2 = Reaction3, event = event3, type = "interval") ~ Days + frailty(Subject, distribution = "gaussian"),
  dist = "gaussian", data = sleepstudy2)

summary(fm.survreg)
```

